Population structure and phenotypic variation of *Sclerotinia sclerotiorum* from dry bean in the United States

### 2017-08-21 doc/manuscript/manuscript.Rmd
Rearrange figures and tables

```{r initiate, echo = FALSE, message = FALSE, warning = FALSE}
library("poppr")
load("../../data/sclerotinia_16_loci.rda")
load("../../data/RDA_results.rda")
dat11cc <- clonecorrect(dat11, ~Region/Source/Host/Year)

@@

vartab <- function(x){
  tot <- x$tot.chi
  con <- x$constr.chi
  uco <- x$unconst.chi
  data.frame(list(Inertia = c(tot, con, uco),
                  Proportion = c(tot, con, uco)/tot))
}
RDA2R2 <- prettify(vegan::RsquareAdj(cap11cc2)$adj.r.squared)/100
RDA1R2 <- prettify(vegan::RsquareAdj(cap11cc)$adj.r.squared)/100

anova1res <- setNames(cap11anova$`Pr(>F)`, rownames(cap11anova))
anova2res <- setNames(cap11anova2$`Pr(>F)`, rownames(cap11anova2))
```

@@

A total of 366 isolates were collected from 2003 to 2012 (except 2006 and 2011) from diseased dry bean plants in Australia, Mexico, France, and 11 states in the USA (Table \@ref(tab:isolate-table)).
Over 50% of the isolates came from four states, MI (62), ND (60), WA (59), NE (47). 
Four regions had fewer than 10 isolates, Australia (6), WI (2), NY (1), ID (1).
We observed 87 MCGs, the most abundant of which ('MCG 5') was represented by 73 isolates over 37 MLGs (Fig \@ref(biggraph)A,C).
After clone-correction on the hierarchy of Region/Source/Host/Year, we were left with 318 isolates representing 165 MLGs (Table \@ref(tab:iatab)).
Analysis of the genotype accumulation curve showed no clear plateau (Supplemental Information), indicating that we would likely obtain more multilocus genotypes if we were to genotype more loci. The number of observed alleles per locus ranged from two to 10 with an average of 6.27 (Table \@ref(tab:locus-stats)).

```{r , echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
loc_columns <- cols(
  Locus = col_character(),
  Range = col_character(),
  Repeats = col_double(),
  allele = col_integer(),
  Hexp = col_double(),
  Evenness = col_double()
)
loc_table <- readr::read_csv("../../results/tables/locus-table.csv", 
                             col_types = loc_columns,
                             col_names = TRUE)
loc_table %>%
  mutate(allele = round(allele, 2)) %>%
  mutate(Repeats = case_when(
    Repeats <= 2 ~ "di-",
    Repeats <= 3 ~ "tri-",
    Repeats <= 4 ~ "tetra-",
    Repeats <= 5 ~ "penta-",
    Repeats <= 6 ~ "hexa-"
  )) %>%
  rename(`*h*` = Hexp) %>%
  rename(`No. alleles` = allele) %>%
  rename(`Repeat Motif` = Repeats) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>%
  huxtable::set_number_format(huxtable::everywhere, 3:4, 0) %>%
  huxtable::set_number_format(huxtable::everywhere, 5:6, 2) %>%
  huxtable::set_align(huxtable::everywhere, 2:6, "right") %>%
  huxtable::set_col_width(c(0.06, 0.07, 0.11, 0.1, 0.05, 0.075)) %>%
  huxtable::print_md(max_width = 61)

Amean <- mean(loc_table$allele)
Hmean <- mean(loc_table$Hexp)
Emean <- mean(loc_table$Evenness)
```

@@

@@

```{r table3, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/mlg_table.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character())) %>%
  rename(`$P_{sex}$` = Psex) %>% 
  knitr::kable(format = "pandoc")
```

@@

![
Network of populations (nodes/circles) and their shared multilocus genotypes (MLG) (edges/lines) genotyped over 11 loci. 
Each node is labeled with **name (number of MLGs shared/number of MLGs total).** 
The shade and area of the nodes are proportional to the number of unique MLGs within the node and the inner nodes are proportional to the number of private MLGs to the region (bottom legend). 
Each edge represents a single MLG where its thickness represents the number of populations that share the MLG and the shade represents the value of $P_{sex}$, or the probability of encountering that MLG from two independent meiotic events.
](../../results/figures/publication/mlg-11.pdf){#community-graph width=100%}

@@

```{r table4, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
suppressPackageStartupMessages(library("tidyverse"))
read_csv("../../results/tables/AMOVA-year.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character()))  %>%
  rename(ps = `Phi statistic`) %>%
  # mutate_if(is.numeric, format, digits = 3) %>%
  mutate(ps = signif(as.numeric(ps), digits = 3)) %>%
  mutate(ps = gsub("0\\.00(\\d{1})(\\d{2})", "\\1.\\2e^-3^", ps)) %>%
  mutate(ps = case_when(P > 0.05 ~ ps, TRUE ~ paste0("**", ps, "**"))) %>%
  rename(`$\\Phi$ statistic` = ps) %>%
  rename(`% variation` = `Percent variation`) %>%
  rename(S.S. = `Sum of Squares`) %>%
  rename(`*P*` = P) %>%
  select(c(1:4, 6, 5)) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>% 
  huxtable::set_col_width(c(0.8, 0.15, 0.15, 0.35, 0.5, 0.15)) %>% 
  huxtable::set_align(huxtable::everywhere, 2:6, "right") %>% 
  huxtable::set_number_format(huxtable::everywhere, 2, 0) %>%
  huxtable::set_number_format(huxtable::everywhere, 6, 3) %>%
  huxtable::set_number_format(huxtable::everywhere, 5, 4) %>%
  huxtable::print_md(max_width = 80)
```

@@

Discriminant Analysis of Principal Components (DAPC) was performed grouping by Region with the first `r dapc_region$n.pca` principal components, representing `r prettify(dapc_region$var)`% of the total variance. 
The first discriminant axis (representing `r paxis(dapc_region)`% of the discriminatory power) separated the centroid for the Mexico isolates from the rest of the data, indicating strong differentiation (Fig. \@ref(DAPC)b). 
The second discriminant axis, representing `r paxis(dapc_region, 2)`% of the discriminatory power, separated the centroid for the California isolates.
The mean population assignment probabilities for all populations with n > 10 showed that only isolates from Mexico, CA, and France had > 50% probabilities of being reassigned to their source populations (Fig. \@ref(DAPC)a). 
DAPC grouping by cultivar used the first 20 principal components, representing 89% of the total variance.
The first two discriminant axes failed to separate any of the cultivars where the mean posterior assignment probabilities were `r passp(dapc_host, "G122")`% (G122), `r passp(dapc_host, "Beryl")`% (Beryl), and `r passp(dapc_host, "Bunsi")`% (Bunsi).
DAPC grouping by Region and Year used the first `r dapc_region_year$n.pca` principal components, representing `r prettify(dapc_region_year$var)`% of the total variance.
The Midwestern populations (NE, MI, CO, ND) did not appear to have any variation across time in contrast to WA, which showed a shift in population structure in the last year of sampling, 2008 (Fig. \@ref(DAPC-RY)).
Further analysis of this population revealed that all 12 isolates in WA circa 2008 originated in a wmn and nine of these genotypes were shared with CA, and three were shared with France (Fig. \@ref(community-graph), \@ref(community-graph16)).

@@

@@

<!-- Conceptualization: SEE. -->
<!-- Formal analysis: BSA SEE. -->
<!-- Funding acquisition: SEE. -->
<!-- Investigation: BSA. -->
<!-- Resources: SEE. -->
<!-- Supervision: BSA. -->
<!-- Visualization: BSA SEE. -->
<!-- Writing - original draft: BSA. -->
<!-- Writing - review & editing: SEE. -->

\newpage
# Supplementary Information {-}

>>>>

```{r initiate, echo = FALSE, message = FALSE, warning = FALSE}
library("poppr")
library("tidyverse")
load("../../data/sclerotinia_16_loci.rda")
load("../../data/RDA_results.rda")
dat11cc <- clonecorrect(dat11, ~Region/Source/Host/Year)
pop_size <- table(pop(dat11))

# Utility Functions -------------------------------------------------------

@@

vartab <- function(x){
  tot <- x$tot.chi
  con <- x$constr.chi
  uco <- x$unconst.chi
  data.frame(list(Inertia = c(tot, con, uco),
                  Proportion = c(tot, con, uco)/tot))
}

# Locus Table -------------------------------------------------------------

loc_columns <- cols(
  Locus = col_character(),
  Range = col_character(),
  Repeats = col_double(),
  allele = col_integer(),
  Hexp = col_double(),
  Evenness = col_double()
)
loc_table <- readr::read_csv("../../results/tables/locus-table.csv", 
                             col_types = loc_columns,
                             col_names = TRUE)
Amean <- mean(loc_table$allele)
Hmean <- mean(loc_table$Hexp)
Emean <- mean(loc_table$Evenness)

# DAPC data ---------------------------------------------------------------

# RDA results -------------------------------------------------------------

RDA2R2    <- prettify(vegan::RsquareAdj(cap11cc2)$adj.r.squared)/100
RDA1R2    <- prettify(vegan::RsquareAdj(cap11cc)$adj.r.squared)/100
anova1res <- setNames(cap11anova$`Pr(>F)`, rownames(cap11anova))
anova2res <- setNames(cap11anova2$`Pr(>F)`, rownames(cap11anova2))
```

@@

A total of 366 isolates were collected from 2003 to 2012 (except 2006 and 2011) from diseased dry bean plants in Australia, Mexico, France, and 11 states in the USA (Table \@ref(tab:isolate-table)).
Over 50% of the isolates came from four states, MI (`r pop_size[["MI"]]`), ND (`r pop_size[["ND"]]`), WA (`r pop_size[["WA"]]`), NE (`r pop_size[["NE"]]`). 
Four regions had fewer than 10 isolates, Australia (`r pop_size[["Australia"]]`), WI (`r pop_size[["WI"]]`), NY (`r pop_size[["NY"]]`), ID (`r pop_size[["ID"]]`).
We observed 87 MCGs, the most abundant of which ('MCG 5') was represented by 73 isolates over 37 MLGs (Fig \@ref(biggraph)A,C).
Analysis of the genotype accumulation curve showed no clear plateau (Supplemental Information), indicating that we would likely obtain more multilocus genotypes if we were to genotype more loci. The number of observed alleles per locus ranged from two to 10 with an average of `r signif(Amean, 3)` (Table \@ref(tab:locus-stats)).
After clone-correction on the hierarchy of Region/Source/Host/Year, we were left with `r nInd(dat11cc)` isolates representing `r nmll(dat11cc)` MLGs (Table \@ref(tab:iatab)).

```{r , echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
loc_table %>%
  mutate(allele = round(allele, 2)) %>%
  mutate(Repeats = case_when(
    Repeats <= 2 ~ "di-",
    Repeats <= 3 ~ "tri-",
    Repeats <= 4 ~ "tetra-",
    Repeats <= 5 ~ "penta-",
    Repeats <= 6 ~ "hexa-"
  )) %>%
  rename(`*h*` = Hexp) %>%
  rename(`No. alleles` = allele) %>%
  rename(`Repeat Motif` = Repeats) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>%
  huxtable::set_number_format(huxtable::everywhere, 3:4, 0) %>%
  huxtable::set_number_format(huxtable::everywhere, 5:6, 2) %>%
  huxtable::set_align(huxtable::everywhere, 2:6, "right") %>%
  huxtable::set_col_width(c(0.06, 0.07, 0.11, 0.1, 0.05, 0.075)) %>%
  huxtable::print_md(max_width = 61)
```

@@

![
Network of populations (nodes/circles) and their shared multilocus genotypes (MLG) (edges/lines) genotyped over 11 loci. 
Each node is labeled with **name (number of MLGs shared/number of MLGs total).** 
The shade and area of the nodes are proportional to the number of unique MLGs within the node and the inner nodes are proportional to the number of private MLGs to the region (bottom legend). 
Each edge represents a single MLG where its thickness represents the number of populations that share the MLG and the shade represents the value of $P_{sex}$, or the probability of encountering that MLG from two independent meiotic events.
](../../results/figures/publication/mlg-11.pdf){#community-graph width=75%}

@@

```{r table3, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
read_csv("../../results/tables/mlg_table.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character())) %>%
  rename(`$P_{sex}$` = Psex) %>% 
  knitr::kable(format = "pandoc")
```

@@

@@

```{r table4, echo = FALSE, warning = FALSE, message = FALSE, results = "asis"}
read_csv("../../results/tables/AMOVA-year.csv", 
         na = "NA", # spaces read in as spaces
         col_types = cols(.default = col_character()))  %>%
  rename(ps = `Phi statistic`) %>%
  # mutate_if(is.numeric, format, digits = 3) %>%
  mutate(ps = signif(as.numeric(ps), digits = 3)) %>%
  mutate(ps = gsub("0\\.00(\\d{1})(\\d{2})", "\\1.\\2e^-3^", ps)) %>%
  mutate(ps = case_when(P > 0.05 ~ ps, TRUE ~ paste0("**", ps, "**"))) %>%
  rename(`$\\Phi$ statistic` = ps) %>%
  rename(`% variation` = `Percent variation`) %>%
  rename(S.S. = `Sum of Squares`) %>%
  rename(`*P*` = P) %>%
  select(c(1:4, 6, 5)) %>%
  huxtable::as_huxtable(add_colnames = TRUE) %>% 
  huxtable::set_col_width(c(0.8, 0.15, 0.15, 0.35, 0.5, 0.15)) %>% 
  huxtable::set_align(huxtable::everywhere, 2:6, "right") %>% 
  huxtable::set_number_format(huxtable::everywhere, 2, 0) %>%
  huxtable::set_number_format(huxtable::everywhere, 6, 3) %>%
  huxtable::set_number_format(huxtable::everywhere, 5, 4) %>%
  huxtable::print_md(max_width = 80)
```

@@

Discriminant Analysis of Principal Components (DAPC) was performed grouping by Region with the first `r dapc_region$n.pca` principal components, representing `r prettify(dapc_region$var)`% of the total variance. 
The first discriminant axis (representing `r paxis(dapc_region)`% of the discriminatory power) separated the centroid for the Mexico isolates from the rest of the data, indicating strong differentiation (Fig. \@ref(DAPC)b). 
The second discriminant axis, representing `r paxis(dapc_region, 2)`% of the discriminatory power, separated the centroid for the California isolates.
The mean population assignment probabilities for all populations with n > 10 showed that only isolates from Mexico, CA, and France had > 50% probabilities of being reassigned to their source populations (Fig. \@ref(DAPC)a).

@@

DAPC grouping by cultivar used the first 20 principal components, representing 89% of the total variance.
The first two discriminant axes (representing `r sum(paxis(dapc_host, 1:2))`% of the discriminatory power) failed to separate any of the cultivars where the mean posterior assignment probabilities were `r passp(dapc_host, "G122")`% (G122), `r passp(dapc_host, "Beryl")`% (Beryl), and `r passp(dapc_host, "Bunsi")`% (Bunsi).
DAPC grouping by Region and Year used the first `r dapc_region_year$n.pca` principal components, representing `r prettify(dapc_region_year$var)`% of the total variance.
The Midwestern populations (NE, MI, CO, ND) did not appear to have any variation across time in contrast to WA, which showed a shift in population structure in the last year of sampling, 2008 (Fig. \@ref(DAPC-RY)).
Further analysis of this population revealed that all 12 isolates in WA circa 2008 originated in a wmn and nine of these genotypes were shared with CA, and three were shared with France (Fig. \@ref(community-graph), \@ref(community-graph16)).

@@

 - **Zhian N. Kamvar** analyzed the data, contributed analysis tools, wrote the paper, prepared figures and tables, reviewed drafts of the paper.

 - **Bimal Sajeewa Amadarasa** ...

 - **Rachana Jhala** ...

 - **Serena McCoy** ...

 - **James Steadman** ...

 - **Sydney E. Everhart** ...

# Supplementary Information {-}
